Microxanox: an R package for simulating an aquatic \(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.

### 2023-06-02 inst/manuscript/Response.md

Editor-in-Chief

SoftwareX

Editor and Reviewer comments:

# Reviewer #1:
> 1.  Please consider reference style: Number the references (numbers in square brackets) in the list in the order in which they appear in the text.

*##Rainer: please correct and then make a response here.*

> The authors in section 3.3 claim several sentences that could confuse readers.
2.1 `one discovers how the steady state of the system responds to the environmental conditions`. There are several concerns that the authors should clarify:

> a)  `The steady state`. Is there a unique steady state? Is there a reported contribution concerning a formal analysis of the model in the sense of stability (stable state existence and uniqueness)? Maybe, authors should refer as `a system's stable state (if it exists)`.

**Response:** We have added the text `The package does not include methods for a formal analysis of the stability of the system and users should take care to assess if steady states are unique, and indeed if a steady state has been achieved.`

**Response:** Since we do not provide methods for formal analysis of stability, we always use "steady state" and have removed all methods of "stable state".

>`b)  `The steady state of the system responds`. It should be noted that the steady state does not respond to the environment. Maybe, the authors refer to the numerical result of a steady state as a function of parameter and initial condition (if a steady state is not unique).

**Response:** Thank you. We have edited the text so that we do not state that the steady state responds to the environment.

> 2.2 `finding steady states that correspond to values of one environmental driver`. The software does not find a steady state. Maybe the authors refer to the software running an open loop of the dynamic system under a set of initial conditions and a subset of parameter values (oxygen diffusivity). Then, the numerical evaluation of steady states could be related to environmental conditions in real systems.

**Response:** we do not understand why the reviewer states that the software does not find a steady state. We are very sorry that we are missing some key point of understanding here.

> 2.3 `Two methods for finding steady states are implemented`. Again, the software does not find a steady state. Maybe the authors refer to two simulations proposed to evaluate an open loop simulation of the dynamic model numerically.

**Response:** we do not understand why the reviewer states that the software does not find a steady state. We are very sorry that we are missing some key point of understanding here.

> 2.4 `The first runs a separate simulation for each combination of starting conditions and oxygen diffusivity`. This sentence is confusing. What the authors mean with `separate simulation` and `starting conditions`. Maybe the authors refer to each initial condition and parameter set value (oxygen diffusivity) as an open loop simulation run of the dynamic system. Are these simulations results focused on obtaining a bifurcation diagram Fig. 3 and 4 from {10.1038/s41467-017-00912-x}? Please clarified this point.

**Response:** We have clarified what is meant by "separate simulation" by changing to "independent simulation", and change "starting conditions" to "initial conditions". We hope that this more standard terminology is clearer. Furthermore, we have added that "This is the method used in the @Bush2017 study and was used to obtain the results in figures 3 and 4 of that article."

> 2.5 `The second runs two simulations, one with step-wise and slowly temporally increasing oxygen diffusivity, and the other with step-wise and slowly decreasing oxygen diffusivity`. This sentence needs to be clarified. Authors refer `temporally` as several changes in the value of oxygen diffusivity during a simulation run time. Maybe the authors mean the following: to explore the parametric sensitivity of the dynamic model under oxygen diffusivity variation, subset values of this parameter are numerically evaluated. Two value vectors of oxygen diffusivity are presented step-wise, gradually increasing and decreasing.

**Response:** We are grateful for this different perspective on how to explain the method and have added it to the manuscript ("Put another way, to explore the sensitivity of steady states of the dynamic model under oxygen diffusivity variation, subset values of this parameter are numerically evaluated. Two value vectors of oxygen diffusivity are presented step-wise, gradually increasing and decreasing.")

> 2.6 `We implemented two methods since there is no definitive best method, and in order to check if results were sensitive to choice of method`. This sentence is confusing.

**Response:** Thank you for pointing out this source of confusion. We change the text to "Two strategies are implemented in order to allow comparison of their results." and moved it to just after we state that two strategies are provided.

> a) `two methods` Maybe the authors mean the following: two numerical strategies to explore a parametric sensitivity numerical evaluation.

**Response:** Thank you for this suggestions. We have changed the text to "Two numerical strategies for finding steady states and their sensitivity to parameters are implemented."

> b) `there is no definitive best method`. Under which criteria? The authors present only two numerical evaluations. Please clarified this point.

**Response:** We have removed this claim since it is not demonstrated to be true and is unimportant.

> c)  `in order to check if results were sensitive to choice of method`. This sentence needs to be clarified. There is no comparison between numerical evaluations in any sense. The numerical simulations or simulation results are not sensitive to choice or method. Indeed, the numerical solution of the dynamical model is sensitive to a set of parameters and initial conditions. Maybe the authors mean the following: to evaluate the performance of the numerical parametric sensitivity analysis of the dynamical model concerning one parametric variation (oxygen diffusivity) under two schemes.2.7 It is not clear why there are two `methods` for numerical parametric sensitivity analysis. The authors should include the arguments of this.

**Response:** We removed this text and now simply state that having two methods allows the comparison of their results.

> 2.8 Please consider rewrite the section 3.3 (see previous 2.1 - 2.7 points) in the sense of Parametric sensitivity analysis, Open loop dynamical behavior, Equilibrium points (see Khalil, H. K. (2002). Nonlinear Systems. Prentice Hall, New Jersey.)

**Response:** we are of the opinion that the with the revisions already made and detailed above, and those below, that the terminology is now more in line with this text, and will be clearly understandable by the target audience (theoretical ecologists) even when the section is not completely rewritten.

> 3.  Please consider the following suggestions:

> Line 5 `types of ecosystem`

Changed to  `types of ecosystems`

>  Line 10 `Permanent link to`

Changed to  `Permanent links to`

>  Line 12 `role in the development of understanding about how ecosystems`

`role in understanding how ecosystems` Changed to `Mathematical models play a key role in the development of ecosystem models and the understanding about how ecosystems work`

>  Line 13: `and how they respond`

Changed to  `and respond`

>  Line 15 `have played a influential role is 16 how ecosystems respond to gradual change`

`have been influential in how ecosystems respond to a gradual change` changed to `One area of ecology in which models have been influential is the understanding of their response to a gradual change in an environmental driver [@Scheffer2001].`

>  Line 16 `is a environmental`

Changed to `is an environmental`

>  Line 17 `affects an ecosystem, but 18 is assumed to not be`

Changed to  `affects an ecosystem. Still, it is assumed not to be`

>  Line 27 `change of the system`

Changed to  `change in the system`

>  Line 28 `to anaerobic`

Changed to  `to an anaerobic`

>  Line 34 `termed the oxygen diffusitivity`

Changed to  `termed oxygen diffusivity`

>  Line 38 `This leaves open the question of if and how biodiversity within these types (i.e. functional groups) of bacteria affects the ecosystem`

Changed to  `This leaves the question of if and how bacteria's biodiversity within these types (i.e., functional groups) affects the ecosystem`

>  Line 47 `It was with this goal in mind that we developed the microxanox package`

Changed to  `With this goal in mind, we developed the microxanox package`

>  Line 58 `conditions, addition`

Changed to  `conditions, the addition`

>  Line 60 `some functions to analyse the results as well as to visualize the results to provide a starting`

`some functions to analyze the results and visualize them to provide a starting` changed to `some functions to analyse the results and visualize these to provide a starting point`

>  Line 71 `for running individual simulations and for running a set of simulations`

Changed to  `for running individual simulations and a set of simulations`

>  Line 76 `the vectors and matrices, and to use matrix mathematics`

Changed to  `the vectors and matrices and used matrix mathematics`

>  Line 80 `code have modular structure so that new functionality can be easily added. E.g. temporally`

Changed to  `code have a modular structure so that new functionality can be easily added. E.g., temporally`

>  Line 87 `to maximise simplicity for the user, and`

Changed to  `to maximize simplicity for the user and`

>  Line 95 `results of the run`

Changed to  `simulation results`

>  Line 98 `parameters and re-running the simulations straightforward. In the following sections we`

`parameters, and reruns the simulations straightforwardly. In the following sections, we` changed to `This promotes reproducibility and makes incremental changes of individual parameters with a consecutive and re-running of the simulations straightforward.`

>  Line 107 `contains among other things the`

Changed to  `contains, among other things, the`

>  Line 109 `their meaning and how they are created and have values set and changed please`

Changed to  `their meaning, and how they are created and have values set and changed, please`

>  Line 113 `an object which is identical to the parameter`

Changed to  `an object identical to the parameter`

>  Line 116 `to run the simulation again from`

Changed to  `to rerun the simulation from`

>  Line 119 `The general approach used to find the stable state of the system`

Changed to  `The general approach to finding the numerical value of a system's stable state (if it exists)`

>  Line ...

>  4.  Please, consider English Editing Service or Proofread Your Paper.

### 2023-06-02 inst/manuscript/manuscript_microxanox.tex
Nearly all done...

 \documentclass[preprint,12pt, a4paper]{elsarticle}

%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

 \journal{SoftwareX} % Sets Journal name

\usepackage{lineno} % add
 \linenumbers % turns line numbering on

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
% \usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

\usepackage[unicode=true]{hyperref}

\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=blue,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother

%% \documentclass[preprint,12pt, a4paper]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

\usepackage{float}
\restylefloat{table}

\journal{SoftwareX}

\begin{document}

\begin{frontmatter}

  \title{Microxanox: an R package for simulating an aquatic
\(MICR\)obial ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
    \author[University of Zürich]{Rainer M Krug%
  %
  \fnref{Corresponding Author}}
   \ead{Rainer.Krug@uzh.ch; Rainer@krugs.de} 
    \author[University of Zürich]{Owen L. Petchey%
  %
  }
   \ead{Owen.Petchey@ieu.uzh.ch} 
      \affiliation[University of Zürich]{Department of Evolutionary
Biology and Environmental Studies, Winterthurerstrasse 190, 8057 Zurich}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{Corresponding Author}
    \fntext[2]{Equal contribution}

  \begin{abstract}
  \(Microxanox\) is an R package to simulate a three functional group
  ecosystem (cyanobacteria, phototrophic sulfur bacteria, and
  sulfate-reducing bacteria) with four chemical substrates (phosphorus,
  oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary
  differential equations. Simulations can be run individually or over a
  parameter range, to find stable states. The model can be implemented
  with different numbers of species per functional group. The package is
  constructed in such a way that the results contain the input parameter
  used, so that a saved results can be loaded again and the simulation
  be repeated. Furthermore, the package framework and code should serve
  as a useful starting point for making simulation models of other types
  of ecosystems.
  \end{abstract}
    \begin{keyword}
    reproducibility \sep regime shift \sep stable state \sep 
    ordinary differential equations
  \end{keyword}

 \end{frontmatter}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
C1 & Current code version & v0.9.0 \\
C2 & Permanent links to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
& & \href{https://zenodo.org/record/6624125}{10.5281/zenodo.6624125} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
C8 & If available Link to developer documentation/manual &
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide} \\
C9 & Support email for questions & Rainer.Krug@uzh.ch;
Owen.Petchey@ieu.uzh \\
\end{longtable}

@@

Mathematical models play a key role in the development of ecosystem
models and the understanding about how ecosystems work and respond to
environmental changes
\citep{Binzer2016a, Chaparro-Pedraza2021, Vasseur2005}. They are also
important for developing hypotheses to test in empirical studies. One
area of ecology in which models have been influential is the
understanding of their response to a gradual change in an environmental
driver \citep{Scheffer2001}. An environmental driver is an environmental
condition that affects an ecosystem. Still, it is assumed not to be
affected by the ecosystem, such as the rate of nutrient input into a
lake.

It is conceivable that an ecosystem state, such as the total biomass of
a particular type of bacteria, may remain unchanged when an
environmental driver changes. It is also possible that the ecosystem
state changes gradually. It is also possible that the ecosystem state
changes abruptly to a new state that is difficult to recover from
\citep{Scheffer2001}. This possibility for abrupt, perhaps catastrophic
changes that are difficult to reverse causes considerable concern
\citep{Collins2021, Northrop2021, Vandermeer2019}.

An example where a gradual change of an environmental variable causes an
abrupt change in the system is the switch from an aerobic (oxygen is
available for metabolism) to an anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. This system has been investigated by
\citet{Bush2017} in a simulation study of a mathematical model. Three
types of microbes occur in the model: cyanobacteria (CB) dominating the
oxic state, and two types of sulfur bacteria that dominate the anoxic
state (sulfate reducing bacteria (SB) and phototrophic sulfur bacteria
(PB)). The model shows that gradual change in the rate at which oxygen
could diffuse into the ecosystem (termed oxygen diffusitivity) could
cause catastrophic changes in the ecosystem state that would be
difficult to reverse.

One feature of the study by \citet{Bush2017} was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves the question of if and how bacteria's
biodiversity within these types (i.e., functional groups) affects the
ecosystem response to environmental change. This limitation is not
specific to the study of \citet{Bush2017}. There are few if any studies
of the effects of biodiversity on abrupt transitions between ecosystem
states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change \citet{Limberger2022}, and to base our work on the
work and model of \citet{Bush2017}. With this goal in mind, we developed
the microxanox package \citep{Krug2022}. The first stage of development
was to write code from scratch (as there was no available code to start
from) and to confirm that this new implementation would reproduce the
previously published results. The resulting reproduction is available as
one of the package vignettes:
\href{https://uzh-peg.r-universe.dev/articles/microxanox/partial-reproduction-Bushetal2017.html}{vignette
Partial reproduction of Bush et al}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on \citet{Bush2017}
(See \citet{Bush2017} for a detailed discussion of the model). At the
core of the simulations is a set of ordinary differential equations
(specified in the function \texttt{bushplus\_dynamic\_model()}, though
this function need not be directly called). There are functions for
running individual simulations and a set of simulations across, for
example, a range of environmental conditions.

@@

The framework used when writing this package aimed to maximise
simplicity for the user and to make it straightforward to reproduce
results (see the supplement \citep{Petchey2022} to \citet{Limberger2022}
for an example of how this is used). As such, all the parameters needed
to run a simulation or find a stable state (i.e.~the final state of the
ecosystem) are contained in a single object (which can easily be created
using included functions). This parameter object is given to a function
that runs the simulations and returns the results. The returned results
object is identical to the parameter object but with an additional slot
named \texttt{results}, which contains the simulation results. Thus the
returned results object contains the simulation conditions (parameters)
as well as the results, and can be used to run the simulation again.
This promotes reproducibility and makes incremental changes of
individual parameters with a consecutive re-running of the simulations
straightforward.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package \citep{Soetaert2010}. The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains,
among other things, the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed, please see the
\emph{User Guide} which accompanies the package or is available at
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

@@

\hypertarget{finding-stable-states}{%
\subsection{Finding Stable States}\label{finding-stable-states}}

The general approach to finding the numerical value of a system's stable
state (if it exists) with a specific parameter set is to run the
simulation for a long time and record the final state. When one does
this across a range of environmental conditions, one discovers how the
steady state of the system responds to the environmental conditions. The
package contains functionality for finding steady states that correspond
to values of one environmental driver, namely the value of oxygen
diffusivity.

Two methods for finding steady states are implemented. The first runs a
separate simulation for each combination of starting conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the \citet{Bush2017} study. The second runs two
simulations, one with step-wise and slowly temporally \emph{increasing}
oxygen diffusivity, and the other with step-wise and slowly
\emph{decreasing} oxygen diffusivity. During this temporal environmental
change, the state of the system is recorded just before change to a new
oxygen diffusivity (we term this the \emph{Temporal method}). We
implemented two methods since there is no definitive best method, and in
order to check if results were sensitive to choice of method.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
% 
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Microxanox: an R package for simulating an aquatic MICRobial ecosystem that can occupy OXic or ANOXic states.},
  pdfkeywords={reproducibility, regime shift, steady state, ordinary
differential equations},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Microxanox: an R package for simulating an aquatic \(MICR\)obial
ecosystem that can occupy \(OX\)ic or \(ANOX\)ic states.}
\author{true \and true}
\date{2023-06-02}

\begin{document}
\maketitle
\begin{abstract}
\(Microxanox\) is an R package to simulate a three functional group
ecosystem (cyanobacteria, phototrophic sulfur bacteria, and
sulfate-reducing bacteria) with four chemical substrates (phosphorus,
oxygen, reduced sulfur, and oxidized sulfur) using a set of ordinary
differential equations. Simulations can be run individually or over a
parameter range, to find steady states. The model can be implemented
with different numbers of species per functional group. The package is
constructed in such a way that the results contain the input parameter
used, so that a saved results can be loaded again and the simulation be
repeated. Furthermore, the package framework and code should serve as a
useful starting point for making simulation models of other types of
ecosystems.
\end{abstract}

@@

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.0483}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4138}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5379}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\textbf{Nr.}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Code metadata description}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Please fill in this column}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
C1 & Current code version & v0.9.0 \\
C2 & Permanent links to code/repository used for this code version &
\url{https://github.com/UZH-PEG/microxanox} \\
& & \href{https://zenodo.org/record/6624125}{10.5281/zenodo.6624125} \\
C3 & Code Ocean compute capsule & \\
C4 & Legal Code License & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
\href{https://cran.r-project.org/index.html}{R} \\
C7 & Compilation requirements, operating environments &
\href{https://cran.r-project.org/index.html}{R (\textgreater= 4.1.0)} \\
& & magrittr \\
& & tibble \\
& & ggplot2 \\
& & patchwork \\
& & grDevices \\
& & stats \\
& & mgcv \\
& & deSolve \\
& & dplyr \\
& & tidyr \\
& & stringr \\
& & multidplyr \\
C8 & If available Link to developer documentation/manual &
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide} \\
C9 & Support email for questions &
\href{mailto:Rainer.Krug@uzh.ch}{\nolinkurl{Rainer.Krug@uzh.ch}};
\href{mailto:Owen.Petchey@ieu.uzh}{\nolinkurl{Owen.Petchey@ieu.uzh}} \\
\end{longtable}

@@

Mathematical models play a key role in the development of ecosystem
models and the understanding about how and why ecosystems change when
the environment changes (\protect\hyperlink{ref-Binzer2016a}{Binzer et
al., 2016};
\protect\hyperlink{ref-Chaparro-Pedraza2021}{Chaparro-Pedraza, 2021};
\protect\hyperlink{ref-Vasseur2005}{Vasseur and McCann, 2005}). They are
also important for developing hypotheses to test in empirical studies.
One area of ecology in which models have been influential is the
understanding of their response to a gradual change in an environmental
driver (\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). An
environmental driver is an environmental condition that affects an
ecosystem. Still, it is assumed not to be affected by the ecosystem,
such as the rate of nutrient input into a lake.

It is conceivable that an ecosystem state, such as the total biomass of
a particular type of bacteria, may remain unchanged when an
environmental driver changes. It is also possible that the ecosystem
state changes gradually. It is also possible that the ecosystem state
changes abruptly to a new state that is difficult to recover from
(\protect\hyperlink{ref-Scheffer2001}{Scheffer et al., 2001}). This
possibility for abrupt, perhaps catastrophic changes that are difficult
to reverse causes considerable concern
(\protect\hyperlink{ref-Collins2021}{Collins et al., 2021};
\protect\hyperlink{ref-Northrop2021}{Northrop et al., 2021};
\protect\hyperlink{ref-Vandermeer2019}{Vandermeer and Perfecto, 2019}).

An example where a gradual change of an environmental variable causes an
abrupt change in the system is the switch from an aerobic (oxygen is
available for metabolism) to an anaerobic (oxygen generally unavailable)
state in a microbial ecosystem. This system has been investigated by
Bush et al. (\protect\hyperlink{ref-Bush2017}{2017}) in a simulation
study of a mathematical model. Three types of microbes occur in the
model: cyanobacteria (CB) dominating the oxic state, and two types of
sulfur bacteria that dominate the anoxic state (sulfate reducing
bacteria (SB) and phototrophic sulfur bacteria (PB)). The model shows
that gradual change in the rate at which oxygen could diffuse into the
ecosystem (termed oxygen diffusitivity) could cause catastrophic changes
in the ecosystem state that would be difficult to reverse.

One feature of the study by Bush et al.
(\protect\hyperlink{ref-Bush2017}{2017}) was limited biodiversity.
Specifically, there was no biodiversity within each of the three types
of bacteria. This leaves the question of if and how bacteria's
biodiversity within these types (i.e., functional groups) affects the
ecosystem response to environmental change. This limitation is not
specific to the study of Bush et al.
(\protect\hyperlink{ref-Bush2017}{2017}). There are few if any studies
of the effects of biodiversity on abrupt transitions between ecosystem
states.

We decided to fill this research gap by making a simulation study of how
within functional group biodiversity affects ecosystem responses to
environmental change Limberger et al.
(\protect\hyperlink{ref-Limberger2022}{2022}), and to base our work on
the work and model of Bush et al.
(\protect\hyperlink{ref-Bush2017}{2017}). With this goal in mind, we
developed the microxanox package (\protect\hyperlink{ref-Krug2022}{Krug
and Petchey, 2022}). The first stage of development was to write code
from scratch (as there was no available code to start from) and to
confirm that this new implementation would reproduce the previously
published results. The resulting reproduction is available as one of the
package vignettes:
\href{https://uzh-peg.r-universe.dev/articles/microxanox/partial-reproduction-Bushetal2017.html}{vignette
Partial reproduction of Bush et al}.

@@

The \emph{microxanox} package is for simulating a three functional group
system (\emph{CB}: cyanobacteria, \emph{PB}: phototrophic sulfur
bacteria, and \emph{SB}: sulfate-reducing bacteria) with four chemical
substrates (\emph{P}: phosphorus, \emph{O}: oxygen, \emph{SR}: reduced
sulfur, \emph{SO}: oxidized sulfur). It includes feedback between
organisms and biogeochemical processes and is based on Bush et al.
(\protect\hyperlink{ref-Bush2017}{2017}) (See Bush et al.
(\protect\hyperlink{ref-Bush2017}{2017}) for a detailed discussion of
the model). At the core of the simulations is a set of ordinary
differential equations (specified in the function
\texttt{bushplus\_dynamic\_model()}, though this function need not be
directly called). There are functions for running individual simulations
and a set of simulations across, for example, a range of environmental
conditions.

@@

The framework used when writing this package aimed to maximise
simplicity for the user and to make it straightforward to reproduce
results (see the supplement (\protect\hyperlink{ref-Petchey2022}{Petchey
et al., 2022}) to Limberger et al.
(\protect\hyperlink{ref-Limberger2022}{2022}) for an example of how this
is used). As such, all the parameters needed to run a simulation or find
a steady state (i.e.~the final state of the ecosystem) are contained in
a single object (which can easily be created using included functions).
This parameter object is given to a function that runs the simulations
and returns the results. The returned results object is identical to the
parameter object but with an additional slot named \texttt{results},
which contains the simulation results. Thus the returned results object
contains the simulation conditions (parameters) as well as the results,
and can be used to run the simulation again. This promotes
reproducibility and makes incremental changes of individual parameters
with a consecutive re-running of the simulations straightforward.

@@

A simulation is run using the \texttt{run\_simulation()} function. In
this function, the ODEs are solved using the function \texttt{ode()} in
the package \emph{deSolve} package
(\protect\hyperlink{ref-Soetaert2010}{Soetaert et al., 2010}). The
\texttt{run\_simulation()} function needs only one argument - an object
as created by the function \texttt{new\_runsim\_parameter()}. The
parameter object returned by \texttt{new\_runsim\_parameter()} contains,
among other things, the \texttt{strain\_parameter} object, which can be
created by the function \texttt{new\_strain\_parameter()}. For a
detailed description of the parameter objects, their meaning and how
they are created and have values set and changed, please see the
\emph{User Guide} which accompanies the package or is available at
\href{https://uzh-peg.r-universe.dev/articles/microxanox/User-guide.html}{User
Guide}.

@@

\hypertarget{finding-steady-states}{%
\subsection{Finding Steady States}\label{finding-steady-states}}

The general approach to finding the numerical value of a system's steady
state (if it exists) with a specific parameter set is to run the
simulation for a long time and record the final state. When one does
this across a range of environmental conditions, one discovers how the
steady state of the system changes with the environmental conditions.
The package contains functionality for finding steady states that
correspond to values of one environmental driver, namely the value of
oxygen diffusivity.

Two numerical strategies for finding steady states and their sensitivity
to parameters are implemented. Two strategies are implemented in order
to allow comparison of their results. The first method runs a
independent simulation for each combination of initial conditions and
oxygen diffusivity (we term this the \emph{Replication method}). This is
the method used in the Bush et al.
(\protect\hyperlink{ref-Bush2017}{2017}) study and was used to obtain
the results in figures 3 and 4 of that article.

The second method runs two simulations, one with step-wise and slowly
temporally \emph{increasing} oxygen diffusivity, and the other with
step-wise and slowly \emph{decreasing} oxygen diffusivity. (Put another
way, to explore the sensitivity of steady states of the dynamic model
under oxygen diffusivity variation, subset values of this parameter are
numerically evaluated. Two value vectors of oxygen diffusivity are
presented step-wise, gradually increasing and decreasing.) During this
temporal environmental change, the state of the system is recorded just
before change to a new oxygen diffusivity (we term this the
\emph{Temporal method}).

The package does not include methods for a formal analysis of the
stability of the system and users should take care to assess if steady
states are unique, and indeed if a steady state has been achieved.
